Relative parabolic factorizations of Chevalley groups

### 2015-08-05 paper.tex
Slightly modified proof of Lemma 13
Various fixes

\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{http://dx.doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat{title}{\usebibmacro{string+doi}{\mkbibemph{#1}}}
\DeclareFieldFormat[article]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\DeclareFieldFormat[inproceedings]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\addbibresource{paper.bib}

@@

Another advantage of parabolic factorizations is that they prove useful in the study of asymptotical properties of Chevalley groups.
The following theorem is the first such application. 
\begin{thm} Let $I$ be an ideal of a commutative ring $R$ such that
\begin{itemize} \item either $\sr(I) = 1$;
 \item or $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$. \end{itemize}
In the latter case assume additionally that Generalized Riemann Hypothesis holds.
Then for any $\Phi$ the width of the congruence subgroup $\G(\Phi, R, I)$ with respect to $z_\alpha$ does not exceed $C_1|\Phi| + C_2$. \end{thm}
\noindent In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).

\begin{thm}
Let $\mathcal{O}_S$ be a Dedekind ring of arithmetic type in some global field $k$ having a real embedding. If $\rk\Phi\geqslant2$, then $\G(\Phi,\mathcal{O}_S,I)$ has finite width with respect to $z_\alpha$.
\end{thm}

@@

The following lemma is a relative version of the classical result sometimes called Chevalley---Matsumoto decomposition.
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be the fundamental representation of $\G_{sc}(\Phi, R)$ with the highest weight $\varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\varpi\ssub{s}}=1$, then 
$g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I)$.
\end{lemma}
\begin{proof}
For $I=R$ the statement of the lemma is contained in~\cite[Theorem~1.3]{St78}). So we can write $g=u_1\cdot g'\cdot u_2$, where $u_1\in\U(\Sigma_s^-,R)$, $u_2\in\U(\Sigma_s^+,R)$ and $g'\in\G(\Delta_s,R)$. Since $g\in\G(\Phi,R,I)$, the vector $gv^+$ is $I$-unimodular. On the other hand, $g'u_2v^+=v^+$, so $u_1\in\G(\Phi,R,I)\cap\U(\Sigma_s^-,R)=\U(\Sigma_s^-,I)$. Now $g'u_2\in\left(\G(\Delta_s,R)\U(\Sigma_s^+,R)\right)\cap\G(\Phi,R,I)$, and by Levi decomposition $u_2\in\U(\Sigma_s^+,I)$, $g'\in\G(\Delta_s,R,I)$.
\end{proof}

@@

\begin{lemma}[{\cite[Corollary~3.3]{S}}]\label{lemma:Stepanov-ideal}
Let $\Phi$ be a root system of rank $\geqslant2$, let $R$ be a commutative ring and $I\trianglelefteq R$ be its ideal.
If $\Phi\neq\rC_\ell$ then $\E\left(\Phi,R,I^2\right)\leqslant\E(\Phi,I)$, otherwise $\E\left(\Phi,R,II^{\indexbox{2}}\right)\leqslant\E(\Phi,I)$.
\end{lemma}
Here $I^{\indexbox{2}}$ stands for the ideal generated by squares $a^2$, where $a\in I$.
Recall also that $I^2$ is generated by products $ab$ for all $a,b\in I$.
Clearly, $II^{\indexbox{2}}$ is generated by elements of the form $a^2b$ for $a,b\in I$.

>>>>

\renewbibmacro{in:}{\ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}
\newbibmacro{string+doi}[1]{\iffieldundef{doi}{\iffieldundef{url}{#1}{\href{\thefield{url}}{#1}}}{\href{http://dx.doi.org/\thefield{doi}}{#1}}}
\DeclareFieldFormat{title}{\usebibmacro{string+doi}{\mkbibemph{#1}}}
\DeclareFieldFormat[article]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\DeclareFieldFormat[inproceedings]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\DeclareFieldFormat[thesis]{title}{\usebibmacro{string+doi}{\mkbibquote{#1}}}
\addbibresource{paper.bib}

@@

Another advantage of parabolic factorizations is that they prove useful in the study of asymptotical properties of Chevalley groups.
The following theorem is the first such application.

Let $G$ be an abstract group with a generating set $X \subset G$. %(for simplicity we assume $X=X^{-1}$ and $1\in X$),
One defines {\it the width of $G$ with respect to $X$} as the smallest natural number $N=w(G,X)$ such that every element of $G$ is a product of at most $N$ elements of $X$.

The relative elementary group $\E(\Phi,R,I)$ is generated by certain elements $z_\alpha(s,\xi)$, see Theorem~\ref{theorem:Tits-Vaserstein}. The width $w(\E(\Phi,R),\mathcal{X})$, where $\mathcal{X}$ is the set of all elementary generators $x_\alpha(\xi)$, has been studied extensively \cite{?}. Denote by $\mathcal{Z}$ the set of all $z_\alpha(s,\xi)$.

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $w(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+3\rk\Phi$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $w(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $w(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
content...
\end{proof}
\noindent In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).

@@

The following lemma is a relative version of the classical result sometimes called Chevalley---Matsumoto decomposition.
\begin{lemma}\label{lemma:Chevalley-Matsumoto}
Let $\pi$ be the fundamental representation of $\G_{sc}(\Phi, R)$ with the highest weight $\varpi_s$.
Assume that $g\in \G_{sc}(\Phi, R, I)$ is such that $(g\cdot v^+)_{\varpi\ssub{s}}=1$, then 
\[ g \in \U(\Sigma_s^-, I) \cdot \G_{sc}(\Delta_s, R, I) \cdot \U(\Sigma_s^+, I). \]
\end{lemma}
\begin{proof}
In the absolute case ($I=R$) the statement of the lemma is contained in~\cite[Theorem~1.3]{St78}. Thus we can write $g=u_1\cdot g'\cdot u_2$, where $u_1\in\U(\Sigma_s^-,R)$, $u_2\in\U(\Sigma_s^+,R)$ and $g'\in\G(\Delta_s,R)$. Since $g$ lies in $\G(\Phi,R,I)$, the vector $gv^+$ is $I$-unimodular. On the other hand, $g'u_2v^+=v^+$, so $u_1\in\G(\Phi,R,I)\cap\U(\Sigma_s^-,R)=\U(\Sigma_s^-,I)$. The matrix $g'u_2$ is an element of $\left(\G(\Delta_s,R)\U(\Sigma_s^+,R)\right)\cap\G(\Phi,R,I)$, and by Levi decomposition $u_2\in\U(\Sigma_s^+,I)$, $g'\in\G(\Delta_s,R,I)$.
\end{proof}

@@

\begin{lemma}[{\cite[Corollary~3.3]{S}}]\label{lemma:Stepanov-ideal}
Let $\Phi$ be a root system of rank $\geqslant2$, let $R$ be a commutative ring and $I\trianglelefteq R$ be its ideal.
If $\Phi\neq\rC_\ell$ then $\E\left(\Phi,R,I^2\right)\leqslant\E(\Phi,I)$, otherwise $\E\left(\Phi,R,II^{\indexbox{2}}\right)\leqslant\E(\Phi,I)$.
\end{lemma}
Here $I^{\indexbox{2}}$ denotes the ideal generated by squares $a^2$, where $a\in I$.
Recall also that $I^2$ is generated by products $ab$ for all $a,b\in I$.
Clearly, $II^{\indexbox{2}}$ is generated by elements of the form $a^2b$ for $a,b\in I$.
